---
title: "How does Social Media relate to Mental Health"
author: "Hoa Ly"
institute: "Dickinson College"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(xaringanthemer)
library(tidyverse)
```
class:center, middle
#Research Question

Can social media behaviors predict mental health symptoms?

---
<style>
.custom-spacing {
  line-height: 1.8;
}
</style>

<div class="custom-spacing">
#Data Overview

- This project use dataset from Kaggle. 
- Dataset consists of: 
  - Age
  - Gender
  - Relationship status
  - Occupation
  - Social media behavior patterns
  - Mental health self-report measures of 481 participants.

<div>
---
<style>
.custom-spacing {
  line-height: 1.8;
}
</style>

<div class="custom-spacing">
#Target Variable

**Mental Health Score** 

Average of: 
- Depressed mood
- Worries
- Sleep issues
- Difficulty concentrating
- Interest fluctuation

**Higher score = worse symptoms**

<div>
---
#Modeling Approach

Data cleaning and processing

80/20 Train-Test split

5-fold cross validation

**Model used:**

1.Multiple Linear Regression

2.Random Forest Regressor

**Evaluation metrics:**
- R^2

- MAE

- RMSE
---
#Model performance

**Multiple Linear Regression**

- Test R^2: 0.402

- MAE: 0.647

- RMSE: 0.794

- CV Mean R²: ~ 0.33

**Random Forest Regression**

- Test R²: 0.405

- MAE: 0.643

- RMSE: 0.792

- CV Mean R²: ~ 0.378

---
#Most important predictors
```{r dev='svg'}
importance_data <- tibble(
  feature = c(
    "Age",
    "Distracted by Social Media",
    "Social Comparison",
    "Restless Without Social Media",
    "Validation Seeking"
  ),
  importance = c(0.195, 0.177, 0.144, 0.098, 0.056)
)

importance_data <- importance_data %>%
  mutate(feature = fct_reorder(feature, importance))

ggplot(importance_data, aes(x = importance, y = feature)) +
  geom_col() +
  labs(
    x = "Feature Importance (Random Forest)",
    y = NULL,
    title = "Top Predictors of Mental Health Score"
  ) +
  theme_minimal(base_size = 20) +
  theme(
    plot.title = element_text(size = 16),
    axis.title.x = element_text(size = 14)
)
```
---
#Key Findings
<style>
.custom-spacing {
  line-height: 1.8;
}
</style>

<div class="custom-spacing">

<div style="font-size: 35px;font-weight: bold;text-align: center;">
"HOW social media is used matters more than HOW MUCH it is used."
</div>

- Compulsive use patterns are strong predictors

- Social comparison strongly associated with worse outcomes

- Platform choice was not among strongest predictors
<div>
---
#Interpratation
<style>
.custom-spacing {
  line-height: 1.8;
}
</style>

<div class="custom-spacing">
Worse mental health scores are associated with:

- Using social media in a distracting way

- Frequently comparing oneself to others

- Seeking validation through likes/comments

- Feeling restless without checking apps

Time spent alone was not the strongest driver.
<div>
---
#Implications for Stakeholders

<style>
.custom-spacing {
  line-height: 1.8;
}
</style>

<div class="custom-spacing">
**Students**

- Focus on reducing comparison and validation cycles

**Universities**

- Promote digital well-being education

**Social Media Platforms**

- Reconsider design features that amplify comparison

**Mental Health Professionals**

- Ask about usage behaviors, not just screen time
<div>
---
#Conclusion


<style>
.custom-spacing {
  line-height: 1.8;
}
</style>

<div class="custom-spacing">
- Social media behaviors moderately predict mental health symptoms

- Behavioral patterns matter more than platform choice

- Compulsive and comparison-driven usage show strongest associations
<div>